Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and neurodevelopment: A study based on the parametric g-formula

### 2024-02-20 output/paper/_figures.qmd
Reviewing manuscript.
Finished reviewing manuscript before sending to MV before sending to co-authors.

  # From table of numerical results to ggplot object
  tbl_num_res <- obj$numerical_results
  if (is_cortisol == FALSE) {
    tbl_num_res <- tbl_num_res |>
    dplyr::left_join(
      edcs[, c("chem_id", "short_name")],
      by = c("variable" = "chem_id")
    ) |>
    dplyr::select(-variable) |>
    dplyr::rename(variable = "short_name") |>
    dplyr::relocate(variable)
  }
  tbl_num_res <- tbl_num_res |>
    tidylog::mutate(
      variable = forcats::fct_reorder2(
        variable, estimate, class
      )
    ) |>
    tidylog::select(-estimate) |>
    ggplot2::ggplot(
      ggplot2::aes(
        y = variable,
        shape = outcome
      )
    ) +
    ggplot2::geom_text(
      ggplot2::aes(
        x = 0,
        label = variable
      ),
      hjust = 0,
      fontface = "bold"
    ) +
    ggplot2::geom_text(
      ggplot2::aes(
        x = 1,
        label = val
      ),
      hjust = 0,
      position = ggstance::position_dodgev(height = 0.9)
    ) +
    ggplot2::theme_void() +
    ggplot2::theme(
      strip.background = ggplot2::element_blank(),
      strip.text = ggplot2::element_blank(),
      legend.position = "bottom",
      legend.box = "vertical",
      legend.margin = ggplot2::margin()
    ) +
    ggplot2::coord_cartesian(
      xlim = c(0, 4)
    )

@@

obj3 <- tidy_res_meffects(
  df = marginal_comparisons_3,
  sa_var = NULL,
  outcome = c("hitrtse"),
  which_res = "comparisons",
  num_digits_est = num_digits_est,
  num_digits_sig = num_digits_sig
)
forest_plot(obj = obj3, r = 5, is_cortisol = TRUE)
```
**Marginal contrasts for the effect of a decrease from the 90th to the 10th percentile of the glucocorticosteroids on {{< acr hitrtse >}} expressed in {{< acr mss >}} (HELIX subcohort; 2013-2016).** 
Circles indicate effect estimates. Solid lines indicate the $95\%$ {{< acr ci >}}.
:::

>>>>

  # From table of numerical results to ggplot object
  tbl_num_res <- obj$numerical_results
  if (is_cortisol == FALSE) {
    tbl_num_res <- tbl_num_res |>
      dplyr::left_join(
        edcs[, c("chem_id", "short_name")],
        by = c("variable" = "chem_id")
      ) |>
      dplyr::select(-variable) |>
      dplyr::rename(variable = "short_name") |>
      dplyr::relocate(variable)
  }
  tbl_num_res <- dplyr::bind_rows(
    tibble::tibble(
      variable = "Exposure",
      val = "Marginal contrast (95% CI)",
      estimate = 100,
      outcome = "",
      class = "AAA"
    ),
    tbl_num_res
  ) |>
    dplyr::group_by(class) |>
    dplyr::arrange(
      variable,
      .by_group = TRUE
    ) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      variable = factor(variable, levels = unique(variable)),
      variable = forcats::fct_rev(variable)
    ) |>
    tidylog::select(-estimate) |>
    ggplot2::ggplot(
      ggplot2::aes(
        y = variable,
        shape = outcome
      )
    ) +
    ggplot2::geom_text(
      ggplot2::aes(
        x = 0,
        label = variable
      ),
      hjust = 0,
      fontface = "bold"
    ) +
    ggplot2::geom_text(
      ggplot2::aes(
        x = 1,
        label = val
      ),
      hjust = 0,
      position = ggstance::position_dodgev(height = 0.9)
    ) +
    ggplot2::theme_void() +
    ggplot2::theme(
      strip.background = ggplot2::element_blank(),
      strip.text = ggplot2::element_blank(),
      legend.position = "bottom",
      legend.box = "vertical",
      legend.margin = ggplot2::margin()
    ) +
    ggplot2::coord_cartesian(
      xlim = c(0, 4)
    )

@@

obj3 <- tidy_res_meffects(
  df = marginal_comparisons_3,
  sa_var = NULL,
  outcome = c("hitrtse"),
  which_res = "comparisons",
  num_digits_est = num_digits_est,
  num_digits_sig = num_digits_sig
)
forest_plot(obj = obj3, r = 5, is_cortisol = TRUE)
```
**Marginal contrasts for the effect of a decrease from the 90th to the 10th percentile of the glucocorticosteroids on {{< acr hitrtse >}} expressed in {{< acr mss >}} (HELIX subcohort; 2013-2016).** 
Circles indicate effect estimates. Solid lines indicate the $95\%$ {{< acr ci >}}. Abbreviations: cortisone production (cortisone prod.); cortisol production (cortisol prod.); corticost. prod. (corticosterone production).
:::

### 2024-02-20 output/paper/_si.qmd
Finished reviewing manuscript before sending to MV before sending to co-authors.

@@

::: {#supptbl-info-chems}
```{r info-chems}
edcs_info <- myphd::edcs_information() |>
  tidylog::select(-c(chem_id, smiles, ctd_id, exposome_explorer_id)) |>
  tidylog::mutate(
    var = janitor::make_clean_names(
      short_name,
      replace = c(`-` = ""),
      case = "none"
    ) |>
      tolower()
  ) |>
  tidylog::relocate(var, .after = short_name)
gt::gt(edcs_info,
       groupname_col = "class") |>
  gt::sub_values(
    values = c("NA"),
    replacement = ""
  ) |>
  gt::cols_label(
    full_name = "Compound",
    short_name = "Symbol",
    var = "Variable name",
    pubchem_cid = "PubChem CID",
    parental_compound = "Parental compound"
  ) |>
  gt::tab_style(
    locations = gt::cells_row_groups(group = dplyr::everything()),
    style = list(gt::cell_text(weight = "bold"))
  )
```
**Information about non-persistent {{< acr edcs >}}, including the full compound name, the standard symbol, the used variable name, the identifier from PubChem, and the parental compound.**
:::

::: {#supptbl-met-desc}
```{r met-desc}
desc_chems_mets$desc_mets
```
**Participants glucocorticosteroids concentrations, by cohort and overall (HELIX subcohort; 2013-2016).**
:::

::: {#supptbl-info-mets}
```{r info-mets}
mets_info <- readODS::read_ods(
  path = here::here("docs", "steroids.ods")
) |>
  tidylog::select(analyte, acronym, type, hmdb_id, cas_number)
gt::gt(mets_info,
       groupname_col = "type") |>
  gt::sub_values(
    values = c("NA"),
    replacement = ""
  ) |>
  gt::cols_label(
    analyte = "Metabolite",
    acronym = "Symbol",
    hmdb_id = "HMDB ID",
    cas_number = "CAS number"
  ) |>
  gt::tab_style(
    locations = gt::cells_row_groups(group = dplyr::everything()),
    style = list(gt::cell_text(weight = "bold"))
  ) |>
  gt::tab_footnote(
    footnote = "Abbreviations: Human Metabolome Database (HMDB); Chemical Abstracts Service (CAS).",
  ) |>
  gt::opt_footnote_marks(
    marks = "letters"
  )
```
**Information about the glucocorticosteroids, including the full metabolite name, the standard symbol, the identifier from the HMDB, and the CAS number.**
:::

::: {#supptbl-lloq-mets}
```{r lloq-mets}
lloq_mets <- load_steroids()$loq |>
  tidylog::mutate(
    Metabolite = stringr::str_replace(
      metabolite,
      "^X", ""
    )
  ) |>
  tidylog::rename(
    LLOQ = "loq"
  ) |>
  tidylog::select(-metabolite) |>
  tidylog::select(Metabolite, LLOQ)
gt::gt(lloq_mets |> dplyr::arrange(Metabolite)) |>
  gt::tab_footnote(
    footnote = "Abbreviations: lower limit of quantification (LLOQ)."
  ) |>
  gt::opt_footnote_marks(
    marks = "letters"
  ) |>
  gt::tab_style(
    locations = gt::cells_column_labels(),
    style = list(gt::cell_text(weight = "bold"))
  )
```
**Lower limits of quantification expressed in ng/ml for the glucocorticosteroids (HELIX subcohort; 2013-2016).**
:::

@@

### Balancing weights

@@

@@

### Balancing weights for effect modification

>>>>

### Information about the endocrine disruptors

::: {#supptbl-info-chems}
```{r info-chems}
edcs_info <- myphd::edcs_information() |>
  tidylog::select(-c(chem_id, smiles, ctd_id, exposome_explorer_id)) |>
  tidylog::mutate(
    var = janitor::make_clean_names(
      short_name,
      replace = c(`-` = ""),
      case = "none"
    ) |>
      tolower()
  ) |>
  tidylog::relocate(var, .after = short_name) |>
  dplyr::group_by(class) |>
  dplyr::arrange(
    short_name,
    .by_group = TRUE
  )
gt::gt(edcs_info,
       groupname_col = "class") |>
  gt::sub_values(
    values = c("NA"),
    replacement = ""
  ) |>
  gt::cols_label(
    full_name = "Compound",
    short_name = "Symbol",
    var = "Variable name",
    pubchem_cid = "PubChem CID",
    parental_compound = "Parental compound"
  ) |>
  gt::tab_style(
    locations = gt::cells_row_groups(group = dplyr::everything()),
    style = list(gt::cell_text(weight = "bold"))
  )
```
**Information about non-persistent {{< acr edcs >}}, including the full compound name, the standard symbol, the used variable name, the identifier from PubChem, and the parental compound.**
:::

### Information about the glucocorticosteroids

::: {#supptbl-info-mets}
```{r info-mets}
mets_info <- readODS::read_ods(
  path = here::here("docs", "steroids.ods")
) |>
  tidylog::select(analyte, acronym, type, hmdb_id, cas_number) |>
  dplyr::group_by(type) |>
  dplyr::arrange(
    analyte,
    .by_group = TRUE
  )
gt::gt(mets_info,
       groupname_col = "type") |>
  gt::sub_values(
    values = c("NA"),
    replacement = ""
  ) |>
  gt::cols_label(
    analyte = "Metabolite",
    acronym = "Symbol",
    hmdb_id = "HMDB ID",
    cas_number = "CAS number"
  ) |>
  gt::tab_style(
    locations = gt::cells_row_groups(group = dplyr::everything()),
    style = list(gt::cell_text(weight = "bold"))
  ) |>
  gt::tab_footnote(
    footnote = "Abbreviations: Human Metabolome Database (HMDB); Chemical Abstracts Service (CAS).",
  ) |>
  gt::opt_footnote_marks(
    marks = "letters"
  )
```
**Information about the glucocorticosteroids, including the full metabolite name, the standard symbol, the identifier from the HMDB, and the CAS number.**
:::

### Codebooks

@@

### Lower limits of quantification of the glucocorticosteroids

::: {#supptbl-lloq-mets}
```{r lloq-mets}
lloq_mets <- load_steroids()$loq |>
  tidylog::mutate(
    Metabolite = stringr::str_replace(
      metabolite,
      "^X", ""
    )
  ) |>
  tidylog::rename(
    LLOQ = "loq"
  ) |>
  tidylog::select(-metabolite) |>
  tidylog::select(Metabolite, LLOQ) |>
  dplyr::arrange(dplyr::desc(LLOQ))
gt::gt(lloq_mets) |>
  gt::tab_footnote(
    footnote = "Abbreviations: lower limit of quantification (LLOQ)."
  ) |>
  gt::opt_footnote_marks(
    marks = "letters"
  ) |>
  gt::tab_style(
    locations = gt::cells_column_labels(),
    style = list(gt::cell_text(weight = "bold"))
  )
```
**Lower limits of quantification expressed in ng/ml for the glucocorticosteroids (HELIX subcohort; 2013-2016).**
:::

### Concentrations of the glucocorticosteroids

::: {#supptbl-met-desc}
```{r met-desc}
targets::tar_load("desc_chems_mets", store = store)
desc_chems_mets$desc_mets
```
**Participants glucocorticosteroids concentrations, by cohort and overall (HELIX subcohort; 2013-2016).**
:::

@@

### Balancing weights: sample sizes

@@

### Balancing weights: summary statistics

@@

### Balancing weights for effect modification: summary statistics

### 2024-02-20 output/paper/_tables.qmd
Finished reviewing manuscript before sending to MV before sending to co-authors.

<!-- title: "Tables" -->

<!-- format: -->
<!--   docx: -->
<!--     reference-doc: reference_document.docx -->

<!-- link-citations: true -->
<!-- linkcolor: blue -->

<!-- editor_options: -->
<!--   chunk_output_type: console -->

<!-- execute: -->
<!--   echo: false -->
<!--   warning: false -->
<!--   error: false -->
<!--   message: false -->

@@

::: {#tbl-pop-desc}
```{r pop-desc}
targets::tar_load("table1", store = store)
table1
  # gt::cols_width(
  #   label ~ gt::px(300)
  # ) |>
  # gt::tab_style(
  #   style = gt::cell_text(size = gt::pct(90)),
  #   locations = gt::cells_body(
  #     columns = dplyr::everything()
  #   )
  # )
```
**Participant characteristics (HELIX subcohort; 2013-2016).**
:::

@@

::: {#tbl-met-new-desc}
```{r met-new-desc}
metabolites <- load_steroids()
scores <- create_steroid_scores(dat = metabolites$metabolome) |>
  tidylog::select(
    HelixID,
    cortisol_production, cortisone_production,
    corticosterone_production
  )
scores <- myphd::extract_cohort(
  dat = scores,
  id_var = "HelixID",
  st = 1, en = 3
) |>
  tidylog::select(-HelixID) |>
  tidylog::mutate(
    cohort = dplyr::case_when(
      cohort == "EDE" ~ "EDEN",
      cohort == "KAN" ~ "KANC",
      cohort == "MOB" ~ "MOBA",
      cohort == "RHE" ~ "RHEA",
      .default = cohort
    )
  ) |>
  tidylog::rename(
    "cortisol production" = "cortisol_production",
    "cortisone production" = "cortisone_production",
    "corticosterone production" = "corticosterone_production"
  )

>>>>

@@

::: {#tbl-pop-desc}
```{r pop-desc}
targets::tar_load("table1", store = store)
table1
```
**Participant characteristics (HELIX subcohort; 2013-2016).**
:::

@@

::: {#tbl-met-new-desc}
```{r met-new-desc}
metabolites <- load_steroids()
scores <- create_steroid_scores(dat = metabolites$metabolome) |>
  tidylog::select(
    HelixID,
    cortisol_production, cortisone_production,
    corticosterone_production
  )
scores <- myphd::extract_cohort(
  dat = scores,
  id_var = "HelixID",
  st = 1, en = 3
) |>
  tidylog::select(-HelixID) |>
  tidylog::mutate(
    cohort = dplyr::case_when(
      cohort == "EDE" ~ "EDEN",
      cohort == "KAN" ~ "KANC",
      cohort == "MOB" ~ "MOBA",
      cohort == "RHE" ~ "RHEA",
      cohort == "SAB" ~ "INMA",
      .default = cohort
    )
  ) |>
  tidylog::rename(
    "cortisol production" = "cortisol_production",
    "cortisone production" = "cortisone_production",
    "corticosterone production" = "corticosterone_production"
  )

### 2024-02-20 output/paper/manuscript.qmd
Finished reviewing manuscript before sending to MV before sending to co-authors.

@@

Levels of unprocessed non-persistent {{< acr edcs >}}, after imputation of values below the {{< acr lod >}}, and glucocorticosteroids, are presented in @tbl-edc-desc, @tbl-met-new-desc, and @supptbl-met-desc.

@@

@Tbl-hypothesis-1and2 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for the {{< acr edcs >}} on the glucocorticosteroids and {{< acr hitrtse >}}. For {{< acr hitrtse >}}, significant differences were present for the phenol OXBE (`r format_res(heffects1, "oxbe", "HRT-SE")`) and the phthalate metabolites MEP (`r format_res(heffects1, "mep", "HRT-SE")`) and MbZP (`r format_res(heffects1, "mbzp", "HRT-SE")`). For the glucocorticosteroids, significant differences were present across all three classes of {{< acr edcs >}} and for all outcomes. The largest differences were attributable to the {{< acr opps >}} DMTP (cortisol production, `r format_res(heffects2, "dmtp", "cortisol production")`) and DETP (corticosterone production, (`r format_res(heffects2, "detp", "corticosterone production")`); cortisone production, (`r format_res(heffects2, "detp", "cortisone production")`)). The forest plots of the individual {{< acr mcs >}} are presented in Supplementary Figures -@suppfig-marginal-1sa and -@suppfig-marginal-2sa.

>>>>

The R code to reproduce analyses and results is available online (https://github.com/lorenzoFabbri/paper-helixSC-neuro).

@@

Levels of unprocessed non-persistent {{< acr edcs >}}, after imputation of values below the {{< acr lod >}}, and glucocorticosteroids, are presented in @tbl-edc-desc, @tbl-met-new-desc, and @supptbl-met-desc. Supplementary Figures [-@suppfig-desc-chems] and [-@suppfig-desc-mets] provide information on the measurement classification of the {{< acr edcs >}} and glucocorticosteroids by cohort, respectively.

@@

@Tbl-hypothesis-1and2 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for the {{< acr edcs >}} on the glucocorticosteroids and {{< acr hitrtse >}}. For {{< acr hitrtse >}}, significant differences were present for the phenol OXBE (`r format_res(heffects1, "oxbe", "HRT-SE")`) and the phthalate metabolites MEP (`r format_res(heffects1, "mep", "HRT-SE")`) and MbZP (`r format_res(heffects1, "mbzp", "HRT-SE")`). For the glucocorticosteroids, significant differences were present across all three classes of {{< acr edcs >}} and for all outcomes. The largest differences were attributable to the {{< acr opps >}} DMTP (cortisol production, `r format_res(heffects2, "dmtp", "cortisol production")`) and DETP (corticosterone production, (`r format_res(heffects2, "detp", "corticosterone production")`); cortisone production, (`r format_res(heffects2, "detp", "cortisone production")`)). The forest plots of the individual {{< acr mcs >}} are presented in Supplementary Figures [-@suppfig-marginal-1sa] and [-@suppfig-marginal-2sa].
